### 2024-06-25 basic-statistics.qmd
20240624 - figures

For a test of whether two variables are associated, the *p*-value evaluates the likelihood that you would observe an association as strong or stronger than the one you observed between the groups if there were no association between the variables.
For instance, when evaluating whether number of carries is <u>positively</u> associated with injury likelihood, and you observed a correlation coefficient of $r = .25$ between number of carries and injury likelihood, the *p*-value evaluates the likelihood that you would observe a correlation as strong or stronger than $r = .25$ between the variables if number of carries is not associated with injury likelihood.

>>>>

For a test of whether two variables are associated, the *p*-value evaluates the likelihood that you would observe an association as strong or stronger than the one you observed between the groups if there were no association between the variables, as depicted in @fig-pValues.
For instance, when evaluating whether number of carries is <u>positively</u> associated with injury likelihood, and you observed a correlation coefficient of $r = .25$ between number of carries and injury likelihood, the *p*-value evaluates the likelihood that you would observe a correlation as strong or stronger than $r = .25$ between the variables if number of carries is not associated with injury likelihood.

```{r}
#| layout-ncol: 2
#| fig-cap: "*p*-Values"
#| fig-subcap: 
#|   - "What is the probability my data would look like this..."
#|   - "...if in the population the association was really this?"
#| label: fig-pValues
#| code-fold: true

set.seed(52242)
correlations <- data.frame(criterion = rnorm(100))
correlations$sample <- complement(correlations$criterion, .95)
correlations$population <- complement(correlations$criterion, 0)

ggplot2::ggplot(
  data = correlations,
  mapping = aes(
    x = sample,
    y = criterion
  )
) +
  geom_point() +
  ggplot2::labs(
    x = "Predictor Variable",
    y = "Outcome Variable",
    title = "What is the probability my data would look like this..."
  ) +
  ggplot2::theme_classic(
    base_size = 16) +
  ggplot2::theme(
    legend.title = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank())

ggplot2::ggplot(
  data = correlations,
  mapping = aes(
    x = population,
    y = criterion
  )
) +
  geom_point() +
  ggplot2::labs(
    x = "Predictor Variable",
    y = "Outcome Variable",
    title = "...if in the population, the association was really this:"
  ) +
  ggplot2::theme_classic(
    base_size = 16) +
  ggplot2::theme(
    legend.title = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank())
```

### 2024-06-25 multiple-regression.qmd
20240605 - getting started/load packages
20240608 - getting started
20240624 - figures

# Factor Analysis {#sec-factorAnalysis}

## Getting Started

### Load Packages

@@

>>>>

# Multiple Regression {#sec-multipleRegression}

## Getting Started {#sec-multipleRegressionGettingStarted}

### Load Packages {#sec-multipleRegressionLoadPackages}

@@

## Overview of Multiple Regression {#sec-multipleRegressionOverview}

Multiple regression examines the association between multiple predictor variables and one outcome variable.
It allows obtaining a more accurate estimate of the unique contribution of a given predictor, by controlling for other variables ([covariates](#sec-covariates)).

## Components {#sec-multipleRegressionComponents}

- $B$ = unstandardized coefficient: direction and magnitude of the estimate (original scale)
- $\beta$ = standardized coefficient: direction and magnitude of the estimate (standard deviation scale)
- $SE$ = standard error: uncertainty of unstandardized estimate

## Coefficient of Determination ($R^2$) {#sec-multipleRegressionRSquared}

::: {#fig-regression}
![](images/multipleRegressionRSquared.png)

Conceptual Depiction of Proportion of Variance Explained ($R^2$) in an Outcome Variable ($Y$) by Multiple Predictors ($X1$ and $X2$) in Multiple Regression. The size of each circle represents the variable's variance. The proportion of variance in $Y$ that is explained by the predictors is depicted by the areas in orange. The dark orange space ($G$) is where multiple predictors explain overlapping variance in the outcome. Overlapping variance that is explained in the outcome ($G$) will not be recovered in the regression coefficients when both predictors are included in the regression model. From @Petersen2024a and @PetersenPrinciplesPsychAssessment.
:::

## Covariates {#sec-covariates}

Covariates are variables that you include in the statistical model to try to control for them so you can better isolate the unique contribution of the predictor variable(s) in relation to the outcome variable.
Use of covariates examines the association between the predictor variable and the outcome variable when holding people's level constant on the covariates.
Inclusion of confounds as covariates allows potentially gaining a more accurate estimate of the causal effect of the predictor variable on the outcome variable.
Ideally, you want to include any and all confounds as covariates.
As described in @sec-correlationCausation, confounds are third variables that influence both the predictor variable and the outcome variable and explain their association.
Covariates are potentially (but not necessarily) confounds.
For instance, you might include the player's age as a covariate in a model that examines whether a player's 40-yard dash time at the NFL Combine predicts their fantasy points in their rookie year, but it may not be a confound.

## Multicollinearity {#sec-multipleRegressionMulticollinearity}

::: {#fig-regression}
![](images/multipleRegressionMulticollinearity.png)

Conceptual Depiction of Multicollinearity in Multiple Regression. From @Petersen2024a and @PetersenPrinciplesPsychAssessment.
:::

### 2024-06-25 research-methods.qmd
20240624 - figures

>>>>

For another example, consider that ice cream sales are associated with shark attacks.
It is unlikely that more people eating ice creams leads to shark attacks.
There is a likely a third variable—heat waves—that is a confound because it influences both ice cream sales and shark attacks and explains their association.
